Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update Obesity dataset

In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response. We utilize the obesity dataset included in the `micemd` package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:

@@

Age and FamOb are missing according to a Missing At Random (MAR) mechanism, whereas the weight variable is subject to selection bias, following an indirect Missing Not At Random (MNAR) mechanism, that usually happens when it is unobserved or omitted a variable affecting both the value of incomplete variable (here weight) and its probability of missing.

The analysis model considers BMI as the dependent variable, predicted by Age, Height, Gender, and FAV. Random slopes are considered for the Age variable, which contains missing values.

The model is represented as:
$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
We load the data:
```{r obesity}
data("data_heckman", package = "micemd")
dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:
```{r obesity-md}
plot_pattern(dat)
```

We observe that the proportion of missingness for the Age and FAVC variables is quite similar across the clusters. However, regarding the weight variable, we notice a heterogeneous proportion of missingness, with systematic missingness in cluster 5.
In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:

```{r obesity icc}
dat$bmi=dat$Weight/(dat$Height)^2
Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
#Performance::ICC(Nulmodel)
```

Since the ICC is above 0.1, we decide to employ two-level (2l) imputation methods. In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable. We also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters. It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age and weight variables. However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.ppm." For the BMI variable, we employ deterministic imputation.
```{r obesity meth}
meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
```
For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables.

After confirming convergence, we proceed to fit the main model and save the results for future use. Users are considering the possibility that the weight variable may not have been selected randomly. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

@@

Create predictor matrix:
```{r obesity-pred}
pred <- quickpred(dat)   # predictor matrix
pred[,"Cluster"]<- -2 # clustering variable
pred[, "Weight"] <- -3 #inclusion-restriction variable
plot_pred(pred)
```

Set the Heckman model as imputation method:
```{r meth}
meth <- make.method(dat) # methods vector
meth["weight"]<-"2l.heckman"
```

Impute the missingness:
```{r eval=FALSE, cache=TRUE}
imp <- mice(
  dat, # dataset with missing values
  m = 5, # number of imputations
  maxit = 10,
  meth = meth, #imputation method vector
  pred = pred, #imputation predictors matrix
  meta_method = "reml",
  printFlag = FALSE
)
```

To apply the Munoz method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package. Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based on certain variables.

TODO: explain exclusion restriction.
In this specific case, it's assumed that height does not influence selection (-4), and an exclusion restriction variable is introduced as an instrumental variable affecting selection but not the outcome. Time of response is used for this purpose, as it's believed that subjects taking more time to complete the survey are more likely to not respond due to poor connectivity or lack of familiarity with electronic surveys (specified as -3).

After implementing these imputation procedures, it's crucial to check for convergence. If convergence issues arise and imputed values are far from plausible, especially affecting BMI values, a post-imputation procedure is employed to constrain BMI values within a plausible range.

Then after this modification we proceed to compare the effects on the model
###Plot estimates

The imputation of these data is based on the [IPDMA Heckman Github repo](https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R)

>>>>

In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response. We use the obesity dataset included in the `micemd`@ package, a synthetic dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:

@@

In this dataset, Age and FamOb are MAR, while the weight variable is affected by selection bias, attributed to an indirect MNAR mechanism. This MNAR mechanism typically arises when an unobserved or omitted variable influences both the value of the incomplete variable (in this case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable, with Age, Gender, and FamOb as predictors. Researchers, based on observable data, are considering the inclusion of random intercept effects, as well as introducing a random slope for the Age variable.

The main model is given by:
$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
We start by loading the data:
```{r, include=FALSE}
library(here)
library(performance)
source(here("heckman","ipdfunction.R"),local =knitr::knit_global())
load(here("heckman","Obesity.rda"))
```

```{r obesity}
#data("data_heckman", package = "micemd")
#dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:
```{r obesity-md}
require(patchwork)
  myplots <- lapply(1:5, function(i) {
         ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
           ggtitle(paste0("Cluster", i))+ 
      theme(legend.position = ifelse(i==1,"bottom","none"))
      })
  myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]

We observe that the missing pattern is quite similar across the clusters. However, regarding the weight variable, we notice that is systematically missing in cluster 3.
In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:

```{r obesity icc}
Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
performance::icc(Nulmodel)
```

Since the ICC is above 0.1 and as the main analyis will be consider a mixed model, we decide to use two-level (2l) imputation methods. In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable. We also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters. It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age variable. However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.pmm." For the BMI variable, we employ deterministic imputation.
```{r obsmar_met}
meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
meth_mar["Age"]<-"2l.2stage.pmm" 
```
For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables. We need to suprime the variable Time as this variable is not especified in the main model and modify also in the prediction matrix the relationship between BMI, weight and height to avoid circular predictions. Then we proced to run the imputation model.

```{r obsmar_pred,echo = T}
pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
pred_mar[,"Cluster"]<- -2 # clustering variable
pred_mar[,"Time"]<-0
pred_mar[pred_mar==1]<-2
pred_mar[c("Height", "Weight"), "BMI"] <- 0
ggmice::plot_pred(pred_mar)
```
```{r obsmar_mice,echo = T,results="hide",include=FALSE}
imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
```

```{r obsmar_plot}
summary(complete(imp_mar,"long")$Weight)
plot(imp_mar)
```
We are also contemplating the utilization of the "pmm" option, given that the imputed values' weights are lower than those of the observable values.

```{r obsmarp_mice,results="hide",include=FALSE}
meth_mar["Weight"]<-"2l.2stage.pmm" 
imp_mar_pmm <- mice(data = Obesity, meth = meth_mar, pred = pred_mar, m=10, seed = 123)
```
```{r obsmarp_plot}
summary(complete(imp_mar_pmm,"long")$Weight)
plot(imp_mar_pmm)
```
After confirming convergence, we proceed to save the results for future use. Users are considering the possibility that the weight variable may not have been selected randomly. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

@@

To apply the heckman method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package. Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based (R) on certain variables.

Here for the outcome equation we consider is the same imputation model that we used for the MAR case (main model). 
$$Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}$$
Regarding the selection equation, we include the same predictors as those in the primary model, as well as a time variable. Here the time variable serves as a restriction exclusion variable specifically explaining the probability of being observed but not affecting the incomplete value (Weight). In this context, we assume that the time a user spends completing the survey serves as a proxy for the barriers they may encounter in survey completion, such as familiarity with the survey content or internet speed. These factors may lead the user to skip specific questions or even the entire survey. Also we assume the time is not have any influence on the weight assigned to the subject. 
$$R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}$$
These two equations are jointly estimated under the assumption that the error terms are interconnected with a bivariate normal distribution. For a more comprehensive understanding of the model and the exclusion restriction, you can refer to (Munoz).

To integrate information from both equations, we must adjust the prediction matrix. The cluster variable remains specified as before (-2). In this imputation method, all the variables present in both the selection and outcome equations are included with a random effect. However, it's essential to distinguish which of these variables appear in each equation.

In this framework, when a variable is shared between both equations, it is denoted as (2). Predictors exclusive to the outcome equation are indicated as (-4), while those exclusive to the selection equation are labeled as (-3). Consequently, the only alteration needed in the predictor matrix pertains to the variable "Time."

```{r obsmnar_pred}
pred_mnar <- pred_mar
pred_mnar["Weight","Time"]<- -3
ggmice::plot_pred(pred_mnar)
```

We also need to modify the method of the weight variable
```{r obsmnar_meth}
meth_mnar <- meth_mar
meth_mnar["Weight"]<- "3l.2stage.heckman"
```

Then we proceed to run the imputation model as before 
Impute the missingness:
```{r obsmnar_mice,results="hide",include=FALSE}
imp_mnar <- mice(
  Obesity, # dataset with missing values
  m = 10, # number of imputations
  maxit = 10,
  meth = meth_mnar, #imputation method vector
  pred = pred_mnar, #imputation predictors matrix
  meta_method = "reml",
  pred_std=FALSE,
  printFlag = FALSE
)
```

After executing these imputation procedures, it is essential to assess convergence and the coherence of the imputed values. Upon examining the weight variable, we noticed that the imputed range falls outside the realm of plausible values (as weight should be positive). Consequently, we are contemplating a different approach, specifically, the use of "pmm," which involves imputing values from donor observations. This approach ensures that the imputed values remain within the range of observable values.

```{r obsmnar_plot}
summary(complete(imp_mnar,"long")$Weight)
plot(imp_mnar)
```

We then run the imputation model but this time using the option of pmm, to assure that weight values are in the range of the observable data, this can be implemented by seeting the pmm parameter as true.
```{r obsmnarp_mice,results="hide",include=FALSE}
imp_mnar_pmm <- mice(
  Obesity, # dataset with missing values
  m = 10, # number of imputations
  maxit = 10,
  meth = meth_mnar, #imputation method vector
  pred = pred_mnar, #imputation predictors matrix
  meta_method = "reml",
  pred_std=FALSE,
  pmm=TRUE,
  printFlag = FALSE
)
```

```{r obsmnarp_plot}
summary(complete(imp_mnar_pmm,"long"))
plot(imp_mnar_pmm)
```

Then after this modification we proceed to compare the effects on the model, therefore we run the analysis model on each of the completed datasets along with the dataset where the incomplete values are removed (CC).

```{r models}
library(ggplot2)
cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
```

We note that there is minimal disparity in the age effect, FamObs, across the various imputation models under consideration. Nonetheless, with respect to Gender, we observe a notable distinction: in the M(C)AR assumption, the effect is significant, whereas under the MNAR assumption, it becomes insignificant. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.

### 2023-10-30 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update Obesity dataset

In this example, we demonstrate a multilevel imputation of an intercept
and slope random model with a continuous response. We utilize the
obesity dataset included in the \texttt{micemd} package, a simulated
dataset that emulates an electronic survey in which individuals are
asked to provide information about their weight and consumption habits
in different countries. To easy the explanations we simulate data for 5
clusters and we reduce the dataset to the following variables:

@@

Age and FamOb are missing according to a Missing At Random (MAR)
mechanism, whereas the weight variable is subject to selection bias,
following an indirect Missing Not At Random (MNAR) mechanism, that
usually happens when it is unobserved or omitted a variable affecting
both the value of incomplete variable (here weight) and its probability
of missing.

The analysis model considers BMI as the dependent variable, predicted by
Age, Height, Gender, and FAV. Random slopes are considered for the Age
variable, which contains missing values.

The model is represented as:
\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
We load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
R> dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat)
\end{CodeInput}

@@

We observe that the proportion of missingness for the Age and FAVC
variables is quite similar across the clusters. However, regarding the
weight variable, we notice a heterogeneous proportion of missingness,
with systematic missingness in cluster 5. In order to evaluate if it is
required a imputation method for 1-level or 2-level we assess the
Intraclass Correlation Coefficient (ICC) for the outcome variable, we
use the ``performance'' package:

\begin{CodeChunk}
\begin{CodeInput}
R> dat$bmi=dat$Weight/(dat$Height)^2
R> Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
R> #Performance::ICC(Nulmodel)
\end{CodeInput}
\end{CodeChunk}

Since the ICC is above 0.1, we decide to employ two-level (2l)
imputation methods. In this imputation process, we include all predictor
variables in the main model and since BMI is a composite variable. We
also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use
the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
package, which suggests an appropriate method for MAR variables based on
the type of variable, number of observations in the cluster, and number
of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
and ``2l.2stage.norm'' for the age and weight variables. However, after
inspecting the age density plot, we consider modifying its method to
``2l.2stage.ppm.'' For the BMI variable, we employ deterministic
imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labeled as -2 and the predictor
variable labeled as 2, encompassing all variables.

After confirming convergence, we proceed to fit the main model and save
the results for future use. Users are considering the possibility that
the weight variable may not have been selected randomly. It's likely
that an omitted variable, like self-esteem, could influence this
selection. For instance, individuals with lower self-esteem might have
higher weight values, impacting their willingness to provide honest
information due to embarrassment.

@@

Create predictor matrix:

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred[,"Cluster"]<- -2 # clustering variable
R> pred[, "Weight"] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-pred-1} \end{center}

@@

Set the Heckman model as imputation method:

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> meth["weight"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

Impute the missingness:

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(
+   dat, # dataset with missing values
+   m = 5, # number of imputations
+   maxit = 10,
+   meth = meth, #imputation method vector
+   pred = pred, #imputation predictors matrix
+   meta_method = "reml",
+   printFlag = FALSE
+ )
\end{CodeInput}
\end{CodeChunk}

To apply the Munoz method, the weight variable should be specified as
``2l.2stage.heckman'' found in the micemd package. Additionally, the
prediction matrix needs modification because this method involves
specifying two equations: one for the outcome, describing the incomplete
variable in terms of partially observed predictors (in this case, all
variables from the main model), and the other for the selection model,
explaining the probability of being observed based on certain variables.

TODO: explain exclusion restriction. In this specific case, it's assumed
that height does not influence selection (-4), and an exclusion
restriction variable is introduced as an instrumental variable affecting
selection but not the outcome. Time of response is used for this
purpose, as it's believed that subjects taking more time to complete the
survey are more likely to not respond due to poor connectivity or lack
of familiarity with electronic surveys (specified as -3).

After implementing these imputation procedures, it's crucial to check
for convergence. If convergence issues arise and imputed values are far
from plausible, especially affecting BMI values, a post-imputation
procedure is employed to constrain BMI values within a plausible range.

Then after this modification we proceed to compare the effects on the
model \#\#\#Plot estimates

The imputation of these data is based on the
\href{https://github.com/johamunoz/Heckman-IPDMA/blob/main/Toy_example.R}{IPDMA
Heckman Github repo}

>>>>

In this example, we demonstrate a multilevel imputation of an intercept
and slope random model with a continuous response. We use the obesity
dataset included in the \texttt{micemd}@ package, a synthetic dataset
that emulates an electronic survey in which individuals are asked to
provide information about their weight and consumption habits in
different countries. To easy the explanations we simulate data for 5
clusters and we reduce the dataset to the following variables:

@@

In this dataset, Age and FamOb are MAR, while the weight variable is
affected by selection bias, attributed to an indirect MNAR mechanism.
This MNAR mechanism typically arises when an unobserved or omitted
variable influences both the value of the incomplete variable (in this
case, Weight) and its likelihood of being missing (denoted as R).

In the primary analysis model, BMI serves as the dependent variable,
with Age, Gender, and FamOb as predictors. Researchers, based on
observable data, are considering the inclusion of random intercept
effects, as well as introducing a random slope for the Age variable.

The main model is given by:
\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FamOb_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
We start by loading the data:

\begin{CodeChunk}
\begin{CodeInput}
R> #data("data_heckman", package = "micemd")
R> #dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(patchwork)
R>   myplots <- lapply(1:5, function(i) {
+          ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+ 
+            ggtitle(paste0("Cluster", i))+ 
+       theme(legend.position = ifelse(i==1,"bottom","none"))
+       })
R>   myplots[[1]]+ myplots[[2]] /myplots[[3]]+ myplots[[4]] /myplots[[5]]
\end{CodeInput}

@@

We observe that the missing pattern is quite similar across the
clusters. However, regarding the weight variable, we notice that is
systematically missing in cluster 3. In order to evaluate if it is
required a imputation method for 1-level or 2-level we assess the
Intraclass Correlation Coefficient (ICC) for the outcome variable, we
use the ``performance'' package:

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> performance::icc(Nulmodel)
\end{CodeInput}
\begin{CodeOutput}
# Intraclass Correlation Coefficient

    Adjusted ICC: 0.342
  Unadjusted ICC: 0.342
\end{CodeOutput}
\end{CodeChunk}

Since the ICC is above 0.1 and as the main analyis will be consider a
mixed model, we decide to use two-level (2l) imputation methods. In this
imputation process, we include all predictor variables in the main model
and since BMI is a composite variable. We also incorporate weight and
height variables.

To determine the best imputation method for these variables, we can use
the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
package, which suggests an appropriate method for MAR variables based on
the type of variable, number of observations in the cluster, and number
of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
and ``2l.2stage.norm'' for the age variable. However, after inspecting
the age density plot, we consider modifying its method to
``2l.2stage.pmm.'' For the BMI variable, we employ deterministic
imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7, ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labeled as -2 and the predictor
variable labeled as 2, encompassing all variables. We need to suprime
the variable Time as this variable is not especified in the main model
and modify also in the prediction matrix the relationship between BMI,
weight and height to avoid circular predictions. Then we proced to run
the imputation model.

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- mice(Obesity, maxit = 0)$pred   # predictor matrix
R> pred_mar[,"Cluster"]<- -2 # clustering variable
R> pred_mar[,"Time"]<-0
R> pred_mar[pred_mar==1]<-2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_pred-1} \end{center}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  2.409  69.703  82.925  82.476  95.310 158.420 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmar_plot-2} \end{center}

We are also contemplating the utilization of the ``pmm'' option, given
that the imputed values' weights are lower than those of the observable
values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  27.50   71.02   83.82   83.76   95.55  135.98 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mar_pmm)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmarp_plot-2} \end{center}

After confirming convergence, we proceed to save the results for future
use. Users are considering the possibility that the weight variable may
not have been selected randomly. It's likely that an omitted variable,
like self-esteem, could influence this selection. For instance,
individuals with lower self-esteem might have higher weight values,
impacting their willingness to provide honest information due to
embarrassment.

@@

To apply the heckman method, the weight variable should be specified as
``2l.2stage.heckman'' found in the micemd package. Additionally, the
prediction matrix needs modification because this method involves
specifying two equations: one for the outcome, describing the incomplete
variable in terms of partially observed predictors (in this case, all
variables from the main model), and the other for the selection model,
explaining the probability of being observed based (R) on certain
variables.

Here for the outcome equation we consider is the same imputation model
that we used for the MAR case (main model).
\[Weight_{ij}= \beta^O_{o} + \beta^O_{1}* Age_{ij} + \beta^O_{2}*FamOb_{ij}+ \beta^O_{3}Gender_{ij} + \epsilon^O_{ij}\]
Regarding the selection equation, we include the same predictors as
those in the primary model, as well as a time variable. Here the time
variable serves as a restriction exclusion variable specifically
explaining the probability of being observed but not affecting the
incomplete value (Weight). In this context, we assume that the time a
user spends completing the survey serves as a proxy for the barriers
they may encounter in survey completion, such as familiarity with the
survey content or internet speed. These factors may lead the user to
skip specific questions or even the entire survey. Also we assume the
time is not have any influence on the weight assigned to the subject.
\[R_{ij}= \beta^S_{o} + \beta^S_{1}* Age_{ij} + \beta^S_{2}*FamOb_{ij}+ \beta^S_{3}Gender_{ij} +\beta^S_{4}Time_{ij}+ \epsilon^S_{ij}\]
These two equations are jointly estimated under the assumption that the
error terms are interconnected with a bivariate normal distribution. For
a more comprehensive understanding of the model and the exclusion
restriction, you can refer to (Munoz).

To integrate information from both equations, we must adjust the
prediction matrix. The cluster variable remains specified as before
(-2). In this imputation method, all the variables present in both the
selection and outcome equations are included with a random effect.
However, it's essential to distinguish which of these variables appear
in each equation.

In this framework, when a variable is shared between both equations, it
is denoted as (2). Predictors exclusive to the outcome equation are
indicated as (-4), while those exclusive to the selection equation are
labeled as (-3). Consequently, the only alteration needed in the
predictor matrix pertains to the variable ``Time.''

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> ggmice::plot_pred(pred_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_pred-1} \end{center}

@@

We also need to modify the method of the weight variable

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mnar <- meth_mar
R> meth_mnar["Weight"]<- "3l.2stage.heckman"
\end{CodeInput}
\end{CodeChunk}

Then we proceed to run the imputation model as before Impute the
missingness:

After executing these imputation procedures, it is essential to assess
convergence and the coherence of the imputed values. Upon examining the
weight variable, we noticed that the imputed range falls outside the
realm of plausible values (as weight should be positive). Consequently,
we are contemplating a different approach, specifically, the use of
``pmm,'' which involves imputing values from donor observations. This
approach ensures that the imputed values remain within the range of
observable values.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  10.26   73.41   86.94   88.27  101.73  186.47 
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnar_plot-2} \end{center}

We then run the imputation model but this time using the option of pmm,
to assure that weight values are in the range of the observable data,
this can be implemented by seeting the pmm parameter as true.

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar_pmm,"long"))
\end{CodeInput}
\begin{CodeOutput}
      .imp           .id          Cluster           Time          Gender     
 Min.   : 1.0   Min.   :   1   Min.   :1.000   Min.   :1.016   Female:10720  
 1st Qu.: 3.0   1st Qu.: 528   1st Qu.:2.000   1st Qu.:3.724   Male  :10390  
 Median : 5.5   Median :1056   Median :3.000   Median :5.151                 
 Mean   : 5.5   Mean   :1056   Mean   :2.868   Mean   :5.211                 
 3rd Qu.: 8.0   3rd Qu.:1584   3rd Qu.:4.000   3rd Qu.:6.698                 
 Max.   :10.0   Max.   :2111   Max.   :5.000   Max.   :9.957                 
     Height      FamOb            Age            Weight            BMI       
 Min.   :1.450   no : 3794   Min.   :14.00   Min.   : 27.50   Min.   : 9.59  
 1st Qu.:1.642   yes:17316   1st Qu.:20.00   1st Qu.: 75.19   1st Qu.:25.84  
 Median :1.701               Median :25.00   Median : 88.88   Median :30.57  
 Mean   :1.702               Mean   :24.82   Mean   : 89.61   Mean   :31.02  
 3rd Qu.:1.760               3rd Qu.:29.00   3rd Qu.:103.78   3rd Qu.:35.89  
 Max.   :1.961               Max.   :47.00   Max.   :135.98   Max.   :63.21  
\end{CodeOutput}
\begin{CodeInput}
R> plot(imp_mnar_pmm)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-1} \end{center}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obsmnarp_plot-2} \end{center}

Then after this modification we proceed to compare the effects on the
model, therefore we run the analysis model on each of the completed
datasets along with the dataset where the incomplete values are removed
(CC).

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggplot2)
R> cc_rs<- with(Obesity[complete.cases(Obesity),],lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
R> plot_models(list_models,mod_name=c("Complete case", "MAR","MAR_pmm", "MNAR","MNAR_S", "MNAR_pmm"))
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}

We note that there is minimal disparity in the age effect, FamObs,
across the various imputation models under consideration. Nonetheless,
with respect to Gender, we observe a notable distinction: in the M(C)AR
assumption, the effect is significant, whereas under the MNAR
assumption, it becomes insignificant. An analysis of the intercept
reveals that, under the MNAR assumption, a higher average BMI is
anticipated compared to the MAR assumption.
